---
title: "Sampler comparison across virtual species (2-D experiment)"
output:
  rmarkdown::html_vignette:
    self_contained: false
    toc: true
vignette: >
  %\VignetteIndexEntry{2d-experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  ## Figures as vector PDFs (cairo_pdf embeds glyphs/lines as vectors).
  dev = "cairo_pdf", fig.width = 7, fig.height = 6, dpi = 300,
  message = FALSE, warning = FALSE
)
```

This vignette reproduces the **experimental setting** of the paper
*Pseudo-absences generation through a Markov chain sampler* (§2.3): four virtual
species (VS) in the leading two principal components of the WorldClim bioclimatic
stack, with three pseudo-absence (PA) samplers benchmarked over `R` independent
Bernoulli realisations. It is the reproducible companion the paper points to at
<https://github.com/danddr/GaussNiche>.

> **Compute note.** Running the samplers (`pa_mcmc` Markov chains, `pa_uniform`
> grid calibration via `optimRes`) is **minutes of work**, not a login-node task.
> On a cluster, generate the full `R = 50` results with the SLURM driver
> (`sbatch sbatch/submit_2d_experiment.sh`) and knit this vignette against them;
> the inline fallback below runs a deliberately small demo so the document still
> knits standalone.

```{r libraries}
library(terra)
library(USE.MCMC)
library(ggplot2)
library(sf)
library(hypervolume)
library(MASS)

## knitr sets the working directory to vignettes/, so the repo root is one up.
source("../virtualSpecies_fn.R")
source("../experiment_plots.R")
```

## The four virtual species

Each VS is a bivariate Gaussian suitability surface in (PC1, PC2), normalised to
peak 1. The design is a $2\times2$ factorial crossing **niche breadth**
(generalist vs specialist) with **niche position** (common vs rare); all share
$\rho = 0$ and a 1:1 presence:pseudo-absence ratio.

```{r species-table, echo = FALSE}
catalogue <- list(
  sp1_generalist_common = list(label = "Generalist · common", mu = c(0, 0), sigma_PC1 = 3, sigma_PC2 = 2),
  sp2_specialist_common = list(label = "Specialist · common", mu = c(0, 0), sigma_PC1 = 1, sigma_PC2 = 0.5),
  sp3_generalist_rare   = list(label = "Generalist · rare",   mu = c(4, 1), sigma_PC1 = 3, sigma_PC2 = 2),
  sp4_specialist_rare   = list(label = "Specialist · rare",   mu = c(4, 1), sigma_PC1 = 1, sigma_PC2 = 0.5))
knitr::kable(
  data.frame(
    Species  = names(catalogue),
    Breadth  = c("generalist", "specialist", "generalist", "specialist"),
    Position = c("common", "common", "rare", "rare"),
    mu       = vapply(catalogue, function(s) sprintf("(%g, %g)", s$mu[1], s$mu[2]), character(1)),
    sigma    = vapply(catalogue, function(s) sprintf("(%g, %g)", s$sigma_PC1, s$sigma_PC2), character(1))),
  row.names = FALSE, caption = "The four virtual species (paper Table 1).")
```

## The three samplers and three diagnostics

We benchmark the same presence set at every realisation with three samplers
sharing the GaussNiche sampler interface:

* **RND** (`pa_random`) — uniform draw across the geographic background.
* **USE** (`pa_uniform`) — uniform across the realised environmental space
  (`USE.MCMC::paSampling`).
* **uniform+** (`pa_mcmc`) — the Markov-chain sampler that extends USE to higher
  dimensions with a Gaussian-mixture presence exclusion.

Per (species, sampler, realisation) we record three metrics, stored in
`result$samplers[[name]]$metrics`: hypervolume **overlap** (presence vs PA cloud;
lower separates better), per-axis range **coverage** (`rel_cov_PC*`; PA PC-range /
background PC-range), and **proportion of true absences** (`prop_true_abs`; PAs on
true-absence cells, i.e. where the Bernoulli draw $y_i = 0$; higher is cleaner).
The plotting module discovers the coverage axes
at runtime, so the same code serves the 2-D experiment here and the k-D extension.

## Load the full results, or run a small demo

```{r load-or-run}
user        <- Sys.getenv("USER")
results_rds <- file.path("/cluster/scratch", user, "GaussNiche", "results2d", "summary_2d_full.rds")
demo_n_real <- 6L   # only used by the standalone fallback

if (file.exists(results_rds)) {
  message("Using precomputed R = 50 results: ", results_rds)
  bundle  <- readRDS(results_rds)
  metrics <- bundle$metrics
  source_note <- sprintf("Precomputed SLURM run (R = %d).", bundle$n_realizations)
} else {
  message("No precomputed results found — running a small inline demo (R = ", demo_n_real, ").")
  envData <- terra::rast(USE.MCMC::Worldclim_tmp, type = "xyz")
  rpc     <- rastPCA(envData, stand = TRUE)
  dt      <- na.omit(as.data.frame(rpc$PCs[[c("PC1", "PC2")]], xy = TRUE))
  bw      <- compute_bandwidth(dt)

  shared <- list(
    dt = dt, envData = envData, rho = 0, max_pres = 300, bgk_prev = 1,
    pa_samplers = list(random = pa_random, uniform = pa_uniform, mcmc = pa_mcmc),
    n_realizations = demo_n_real, seed_base = 42, seed_pseudo_base = 123,
    bw = bw, pa_env_rast = envData, verbose = FALSE, parallel = FALSE,
    grid.res = 5, thres = 0.75, chain.length = 4000, burnIn = 500,
    species.cutoff.threshold = 0.1)
  res_list <- lapply(catalogue, function(s)
    do.call(virtualSpecies, c(shared, list(mu = s$mu, sigma_PC1 = s$sigma_PC1, sigma_PC2 = s$sigma_PC2))))
  metrics <- collect_experiment_metrics(
    res_list, species_labels = vapply(catalogue, `[[`, character(1), "label"))
  source_note <- sprintf("Standalone inline demo (R = %d, reduced chain).", demo_n_real)
}
```

## The demanded figure: boxplots of the three metrics

Each box summarises the empirical distribution of a metric across the `R`
Bernoulli realisations; rows are metrics (coverage split per PC axis), columns are
species, and the three boxes per panel are the samplers.

```{r boxplots, fig.height = 7}
plot_experiment_boxplots(metrics, title = "Sampler comparison across virtual species")
```

`r source_note`

### Each metric on its own

```{r per-metric-overlap, fig.height = 3.2}
plot_experiment_metric(metrics, "overlap")
```

```{r per-metric-coverage, fig.height = 4.2}
plot_experiment_metric(metrics, "coverage")
```

```{r per-metric-trueabs, fig.height = 3.2}
plot_experiment_metric(metrics, "prop_true_abs")
```

## Reproducing the full run

The headline figures in the paper come from the full `R = 50` analysis. On the
cluster:

```sh
sbatch sbatch/submit_2d_experiment.sh          # MODE=full, R from N_REALIZATIONS (default 50)
# smoke check first:
sbatch --export=ALL,MODE=smoke --cpus-per-task=2 --time=00:20:00 sbatch/submit_2d_experiment.sh
```

The driver writes per-species `.rds`, a combined `metrics_2d_full.csv`, the
`summary_2d_full.rds` this vignette loads, and the boxplot figures (vector PDF +
PNG) to `<scratch>/GaussNiche/results2d/`.
