---
title: "Sampler comparison across virtual species (2-D and 5-D)"
output:
  rmarkdown::html_vignette:
    self_contained: false
    toc: true
vignette: >
  %\VignetteIndexEntry{2d-experiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  ## Figures as vector PDFs (cairo_pdf embeds glyphs/lines as vectors).
  dev = "cairo_pdf", fig.width = 7, fig.height = 6, dpi = 300,
  message = FALSE, warning = FALSE
)
```

This vignette reproduces the **experimental setting** of the paper
*Pseudo-absences generation through a Markov chain sampler* (§2.3): four virtual
species (VS) in a $2\times2$ factorial (breadth $\times$ position), with several
pseudo-absence (PA) samplers benchmarked over `R` independent Bernoulli
realisations. It is run in **two environmental spaces**: a 2-D space (the leading
two PCs of the WorldClim bioclimatic stack) and a 5-D space (the higher-dimensional
GaussNiche extension). It is the reproducible companion the paper points to at
<https://github.com/danddr/GaussNiche>.

> **Compute note.** Running the samplers (`pa_mcmc` Markov chains, `pa_uniform`
> grid calibration via `optimRes`, the 5-D hypervolumes) is **minutes of work**,
> not a login-node task. On a cluster, generate the full `R = 50` results with the
> SLURM drivers (`sbatch sbatch/submit_2d_experiment.sh`,
> `sbatch sbatch/submit_5d_experiment.sh`) and knit this vignette against them; the
> 2-D inline fallback below runs a deliberately small demo so the document still
> knits standalone. The 5-D section loads precomputed results (it needs the
> `env5d` stack on scratch).

```{r libraries}
library(terra)
library(USE.MCMC)
library(ggplot2)
library(sf)
library(hypervolume)
library(MASS)

## knitr sets the working directory to vignettes/, so the repo root is one up.
source("../virtualSpecies_fn.R")    # 2-D engine + pa_random / pa_uniform / pa_mcmc
source("../experiment_plots.R")     # dimension-agnostic boxplots + report-style figures
source("../pa_buffer.R")            # geographic buffer-out sampler (50 km)
```

## The four virtual species

Each VS is a Gaussian suitability surface, normalised to peak 1. The design crosses
**niche breadth** (generalist vs specialist) with **niche position** (common vs
rare); all share $\rho = 0$ and a 1:1 presence:pseudo-absence ratio. In 2-D the
optimum $\mu$ and breadth $\sigma$ are fixed values; in 5-D they are multiples of
each PC's background standard deviation (see the 5-D section).

```{r species-table, echo = FALSE}
catalogue <- list(
  sp1_generalist_common = list(label = "Generalist · common", mu = c(0, 0), sigma_PC1 = 3, sigma_PC2 = 2),
  sp2_specialist_common = list(label = "Specialist · common", mu = c(0, 0), sigma_PC1 = 1, sigma_PC2 = 0.5),
  sp3_generalist_rare   = list(label = "Generalist · rare",   mu = c(4, 1), sigma_PC1 = 3, sigma_PC2 = 2),
  sp4_specialist_rare   = list(label = "Specialist · rare",   mu = c(4, 1), sigma_PC1 = 1, sigma_PC2 = 0.5))
knitr::kable(
  data.frame(
    Species  = names(catalogue),
    Breadth  = c("generalist", "specialist", "generalist", "specialist"),
    Position = c("common", "common", "rare", "rare"),
    mu       = vapply(catalogue, function(s) sprintf("(%g, %g)", s$mu[1], s$mu[2]), character(1)),
    sigma    = vapply(catalogue, function(s) sprintf("(%g, %g)", s$sigma_PC1, s$sigma_PC2), character(1))),
  row.names = FALSE, caption = "The four virtual species in 2-D (paper Table 1).")
```

## The samplers and the three diagnostics

We benchmark the same presence set at every realisation with samplers sharing the
GaussNiche sampler interface:

* **RND** (`pa_random`) — uniform draw across the geographic background.
* **buffer-out** (`pa_buffer_out`) — uniform draw, but only from cells beyond a
  50 km buffer around the presences (a non-random geographic baseline, cf. the
  original USE paper's primary source).
* **USE** (`pa_uniform`) — uniform across the realised environmental space
  (`USE.MCMC::paSampling`). Grid-based, so **2-D only**.
* **uniform+** (`pa_mcmc`) — the Markov-chain sampler that extends USE to higher
  dimensions with a Gaussian-mixture presence exclusion.

Per (species, sampler, realisation) we record three metrics, stored in
`result$samplers[[name]]$metrics`: hypervolume **overlap** (presence vs PA cloud;
lower separates better), per-axis range **coverage** (`rel_cov_PC*`), and
**proportion of true absences** (`prop_true_abs`). The plotting module discovers
the coverage axes at runtime, so the same code serves 2-D and 5-D.

# 2-D experiment

```{r load-or-run-2d}
user        <- Sys.getenv("USER")
results_2d  <- file.path("/cluster/scratch", user, "GaussNiche", "results2d", "summary_2d_full.rds")
demo_n_real <- 6L   # only used by the standalone fallback

if (file.exists(results_2d)) {
  message("Using precomputed 2-D results: ", results_2d)
  metrics_2d <- readRDS(results_2d)$metrics
  src_2d <- "Precomputed SLURM run (R = 50)."
} else {
  message("No precomputed 2-D results — running a small inline demo (R = ", demo_n_real, ").")
  envData <- terra::rast(USE.MCMC::Worldclim_tmp, type = "xyz")
  rpc     <- rastPCA(envData, stand = TRUE)
  dt      <- na.omit(as.data.frame(rpc$PCs[[c("PC1", "PC2")]], xy = TRUE))
  bw      <- compute_bandwidth(dt)
  shared  <- list(
    dt = dt, envData = envData, rho = 0, max_pres = 300, bgk_prev = 1,
    pa_samplers = list(random = pa_random, buffer = pa_buffer_out,
                       uniform = pa_uniform, mcmc = pa_mcmc),
    n_realizations = demo_n_real, seed_base = 42, seed_pseudo_base = 123,
    bw = bw, pa_env_rast = envData, verbose = FALSE, parallel = FALSE,
    grid.res = 5, thres = 0.75, buffer_km = 50, chain.length = 4000, burnIn = 500,
    species.cutoff.threshold = 0.75)
  res_2d  <- lapply(catalogue, function(s)
    do.call(virtualSpecies, c(shared, list(mu = s$mu, sigma_PC1 = s$sigma_PC1, sigma_PC2 = s$sigma_PC2))))
  metrics_2d <- collect_experiment_metrics(res_2d, species_labels = vapply(catalogue, `[[`, character(1), "label"))
  src_2d <- sprintf("Standalone inline demo (R = %d, reduced chain).", demo_n_real)
}
```

Each box summarises a metric's distribution across the `R` Bernoulli realisations;
rows are metrics (coverage split per PC axis), columns are species, boxes are
samplers.

```{r boxplots-2d, fig.height = 7}
plot_experiment_boxplots(metrics_2d, title = "Sampler comparison across virtual species (2-D)")
```

`r src_2d` Here USE and uniform+ behave alike (uniform+ is a drop-in replacement
for the grid-based USE) while both cover the environmental range better than the
geographic samplers (RND, buffer-out).

# 5-D (high-dimensional) experiment

The same four species are reproduced in a **five-dimensional** environmental space,
following the GaussNiche higher-dimensional extension. The 5-D space augments the
WorldClim climate baseline with climate-orthogonal predictors (soil pH, clay, bulk
density and organic carbon; terrain ruggedness and topographic position; tree
cover) on the same Central/Western-European 10 arc-min grid, keeping the leading
**five principal components (~82.5% of variance)**. The niche optimum and breadth
are now length-5 vectors specified relative to each component's background standard
deviation: **generalist** $\sigma = 1.0\,\mathrm{sd}$, **specialist**
$\sigma = 0.5\,\mathrm{sd}$; the **common** optimum is the mode of the 5-D
background density and the **rare** optimum a fixed peripheral point.

Two things change in 5-D. First, the grid-based **USE sampler is unavailable**
(it is restricted to two dimensions), so the comparison is **RND vs buffer-out vs
uniform+**, with buffer-out as the non-random baseline. Second, *uniform+* is the
only sampler that keeps the pseudo-absences uniform across the *full* environment:
the curse of dimensionality makes the geographic samplers under-cover the trailing
components.

```{r load-5d}
results_5d <- file.path("/cluster/scratch", user, "GaussNiche", "results5d_experiment", "summary_5d_full.rds")
have_5d <- file.exists(results_5d)
if (have_5d) {
  bundle_5d  <- readRDS(results_5d)
  metrics_5d <- bundle_5d$metrics
  message("Using precomputed 5-D results: ", results_5d)
} else {
  message("No precomputed 5-D results found at ", results_5d,
          " — run `sbatch sbatch/submit_5d_experiment.sh` first; the 5-D figures are skipped.")
}
```

```{r boxplots-5d, fig.height = 9, eval = have_5d}
plot_experiment_boxplots(metrics_5d, title = "Sampler comparison across virtual species (5-D)")
```

```{r note-5d, echo = FALSE, results = "asis", eval = !have_5d}
cat("> *5-D results not available in this build; the boxplots above are skipped.*\n")
```

In 5-D, **uniform+ attains the highest range coverage on every principal
component for all four species**, and the advantage widens on the trailing
components (PC3--PC5), which the geographic samplers leave under-sampled.
buffer-out separates strongly (low overlap, high true-absence) but covers the
environment poorly — exactly the geographic-method limitation that *uniform+*
overcomes while needing no grid.

Two further per-species figures are produced for the appendix by the same driver
(`run_5d_experiment.R` via `experiment_plots.R`): a `plot_pc_matrix()` scatterplot
matrix of the niche and pseudo-absences across all PC pairs (lower triangle = RND,
upper = uniform+), and a `plot_geo_sampler_grid()` species $\times$ sampler map of
the pseudo-absences back-projected onto the suitability surface.

# Reproducing the full runs

The headline `R = 50` results come from the SLURM drivers:

```sh
# 2-D (4 samplers; cutoff 0.75):
sbatch sbatch/submit_2d_experiment.sh
# 5-D (RND / buffer-out / uniform+; env5d; cutoff 0.7):
sbatch sbatch/submit_5d_experiment.sh
# fast re-render of all 5-D figures from saved results (no recompute):
sbatch --export=ALL,MODE=figures sbatch/submit_5d_experiment.sh
```

Each driver writes per-species `.rds`, a combined `metrics_*.csv`, the
`summary_*.rds` this vignette loads, and the figures (vector PDF + PNG). The
paper's Results figures, the appendix PC matrices, and the geographic grid are all
produced here and copied into `../markov-chain-sampler-paper/graphics/`.
```
